---
title: "Week 3 Assignment"
author: "Rachel Smith"
date: "3/3/2021"
output: ioslides_presentation

---

Overview
========================================================

- This presentation is in for the week 3 assignment of the Developing Data Products course as part of Coursera JHU Data Science Specialization
- Task: create a web page presentation that features a plot created with Plotly

## Load libraries & data

```{r, message = FALSE, warning = FALSE}

library(tidyverse)
library(plotly)

data(diamonds)

```

Here are the available variables to look at:
```{r vars}

colnames(diamonds)

```

Subset the data
```{r subset}

df_diamonds <- diamonds[sample(nrow(diamonds), 1000),]

```

## Plot 1: Distribution of Diamond Prices

```{r plot 1, echo = FALSE, message = FALSE, warning = FALSE}

p1 <- df_diamonds %>% 
       
       ggplot(aes(x = price)) +
       geom_histogram() 

ggplotly(p1)

```

## Plot 2: Diamond price by carat

```{r plot2, echo = FALSE, message = FALSE, warning = FALSE}

p2 <- df_diamonds %>% 
       
       ggplot(aes(x = carat, y = price)) +
       geom_point(aes(color = clarity)) 

ggplotly(p2)

```





